---
title: "Cardiac EDA"
author: "Michiel Noback (NOMI)"
date: "11/05/2020"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

# Assignment Introduction

The purpose of this assignment is to train you in performing reproducible research.
The remainder of this document logs an exploratory data analysis. It is your task to study this EDA and describe weaknesses with respect to reproducibility.

Reproducibility is guaranteed by the following aspects of an analysis:

1. There is a log of all steps of the analysis  
2. The log is in English and is easy to read (spelling/grammar/formulations)
3. The data are described in detail. It is known  
    - where they came from and how they were collected
    - what the variables are and what abbreviations mean
    - what the data types of the variables are, what values they may have, and in what units they were measured
    - what the dependent / class variable is
4. There are no data processing steps missing in the log (or its rendered result!)  
5. It is clear what the sequence of processing or analysis steps is (also chronologically), and why they were undertaken in that particular order
6. Every step has  
    - an intro: why is it carried out   
    - a result: presented in clear tables or figures or other relevant means  
    - a conclusion: was the step succesful, are the results as expected, what action should additionaly be done, what questions do arise as a result etc.


Here follows the actual EDA. It is written entirely using the _tidyverse_ packages as well as some others. Study it in pairs or trios and report flaws, errors, weaknesses and possible soluitons/repairs at the end of the session.


# EDA of cardiac data: dobutamine efficacy

setop and load libraries:

```{r setup, include=FALSE}
options(digits = 3)
library(dplyr)
library(ggplot2)
library(tidyr)
library(readr)
library(gridExtra)
library(GGally)
```


Load the codebook

```{r load-codebook, message=FALSE}
codebook <- read_delim(file = "codebook.txt",
                       delim = ";")

knitr::kable(codebook)
```

Load the data.

```{r load-data, message=FALSE}
cardiac <- read_csv(file = "cardiac.csv")
spec(cardiac)
```

Let's drop columns 33-42

```{r drop-columns}
cardiac <- cardiac %>% select(1:32)
```

The dimensions are supposed to be 32 columns and 558 rows. Check:

```{r}
cardiac <- as_tibble(cardiac)
cardiac
```


Have a look at the structure of the df:

```{r}
glimpse(cardiac)
```

All colums are numeric. Some however should be factors. That will be dealt with later.

```{r}
summary(cardiac)
```

Some variables were selected for boxplot:

```{r}
my_plots <- list()
#use indices is important!
for (i in 1:6) {
    n <- names(cardiac)[i]
    #use aes_string() !!!
    g <- ggplot(data = cardiac, mapping = aes_string(y = n)) +
        geom_boxplot() +
        ylab(n) +
        ggtitle(n)
    my_plots[[i]] <- g ##has to be integer, not name!
}
#use do.call() to process the list in grid.arrange
do.call(grid.arrange, c(my_plots, nrow = 3))
```

The `bhr` variable has a maximum of 210 which is highly unlikely. Let's investigate the distribution of it:

```{r}
ggplot(data = cardiac, mapping = aes(x = bhr)) +
    geom_histogram(bins = 50)
```

The `210` value is absurd and wrong for sure. I am going to take it out.

```{r}
cardiac <- cardiac %>%
    filter(bhr != max(bhr)) 
```

### Data transformations

The gender attribute will be converted into a factor for easy and readable analysis.

```{r}
cardiac <- cardiac %>% 
    mutate(gender = factor(gender, labels = c("m", "f"), levels = c(0, 1)),
           chestpain = recode(chestpain, "0" = "yes", "1" = "no"),
           any.event = factor(`any event`, labels = c("yes", "no"), levels = c(0, 1)))

# 0 MEANS THE PATIENT EXPERIENCED CHEST PAIN
head(cardiac)
```

## Pairwise plot

```{r warning=FALSE, message=FALSE}
cardiac %>% 
  select(c(1:6, 33)) %>%
  ggpairs(mapping = aes(color = any.event, alpha = 0.3))
```


## Dose dependency of variables

A custom function ripped from the internet (https://stackoverflow.com/questions/3735286/create-a-matrix-of-scatterplots-pairs-equivalent-in-ggplot2)

```{r}
gatherpairs <- function(data,
                        ...,
                        xkey = '.xkey',
                        xvalue = '.xvalue',
                        ykey = '.ykey',
                        yvalue = '.yvalue',
                        na.rm = FALSE,
                        convert = FALSE,
                        factor_key = FALSE) {
  vars <- quos(...)
  xkey <- enquo(xkey)
  xvalue <- enquo(xvalue)
  ykey <- enquo(ykey)
  yvalue <- enquo(yvalue)
  
  data %>% {
    cbind(
      gather(
        .,
        key = !!xkey,
        value = !!xvalue,
        !!!vars,
        na.rm = na.rm,
        convert = convert,
        factor_key = factor_key
      ),
      select(.,!!!vars)
    )
  } %>% gather(
    .,
    key = !!ykey,
    value = !!yvalue,
    !!!vars,
    na.rm = na.rm,
    convert = convert,
    factor_key = factor_key
  )
}
```

Usage:

```{r, out.width="100%", fig.asp=1.5, warning=F}
cardiac %>% 
  mutate(dose = factor(dose)) %>%
  drop_na(c(bhr, basebp, pkhr, maxhr)) %>%
  gatherpairs(bhr, basebp, pkhr, maxhr) %>% {
    ggplot(., aes(x = .xvalue, y = .yvalue, color = dose)) +
      geom_point(size = 0.2, alpha = 0.3) +
      geom_smooth(method = 'lm', size = 0.5) +
      facet_wrap(
        .xkey ~ .ykey,
        ncol = length(unique(.$.ykey)),
        scales = 'free',
        labeller = label_both
      ) +
      scale_color_brewer(type = 'qual')
  }

```

## Relationships with outcome variable


## Different outcome variables

Dose : DOSE OF DOBUTAMINE GIVEN

Which variable is interesting?

Maybe maxhr?

```{r}
cardiac %>%
  ggplot(mapping = aes(x = dose, y = maxhr)) +
    geom_jitter(width = 0.2, alpha = 0.3)
```


```{r}
cardiac %>%
  mutate(dose = factor(dose)) %>%
  ggplot(mapping = aes(x = dose, y = maxhr, color = dose)) +
    geom_boxplot(varwidth = TRUE) + theme(legend.position = "none")

```


```{r}
cardiac %>%
  mutate(dose = factor(dose)) %>%
  ggplot(mapping = aes(x = dose, y = maxhr)) +
    geom_violin(aes(color = dose))

```




# PCA

Does Principal Components Analysis tell me anything about patterns of variation?


```{r}
library(devtools)
#install_github("vqv/ggbiplot")
library(ggbiplot)

#reload data
cardiac <- read.csv(file = "cardiac.csv")
cardiac <- cardiac %>% select(1:32)

cardiac.pca <- prcomp(cardiac[, 1:17], center = TRUE, scale. = TRUE)
summary(cardiac.pca)
ggbiplot(cardiac.pca, ellipse=TRUE, groups = factor(cardiac$any.event), alpha = 0.5, size = 0.2)
```


Boxplots

```{r}
my_plots <- list()
#use indices is important!
for (i in 1:6) {
    n <- names(cardiac)[i]
    #use aes_string() !!!
    g <- ggplot(data = cardiac, mapping = aes_string(y = n)) +
        geom_boxplot() +
        ylab(n) +
        ggtitle(n)
    my_plots[[i]] <- g ##has to be integer, not name!
}
#use do.call() to process the list in grid.arrange
do.call(grid.arrange, c(my_plots, nrow = 3))
```


